---
title: "Bayesian Data Analysis for Medical Data"
subtitle: Bayesian Inference
author      : Paola Berchialla
job         : Department of Clinical and Biological Sciences, University of Torino
output:
  ioslides_presentation:
    css: assets/css/ioslides.css
    logo: assets/img/dscb.jpg
    mathjax: local
    self_contained: no
    widescreen: yes
  beamer_presentation: default
  slidy_presentation: default
---

## Quiz!

```{r quiz, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center'}
library(maps)
kc8084 <- read.delim('data\\gd80to84.txt')
names(kc8084) <- names(kc8084)[-grep('x', tolower(names(kc8084)))]

kc8589 <- read.delim('data\\gd85to89.txt')

kc <- data.frame(state=kc8589$state, 
                 county=kc8589$county, 
                 aakc=kc8084$aadc + kc8589$aadc
                 #aakc=10^5*(kc8084$dc + kc8589$dc)/(kc8084$pop + kc8589$pop)
)

q90 <- with(kc, quantile(aakc, probs = 0.9, na.rm=T))
kc90 <- subset(kc, aakc>q90)

counties.90 <- paste0(tolower(kc90$state),',',tolower(kc90$county))


library(maps)
map('state')
map('county', counties.90, col='black', fill=T, add=T)
title('Highest kidney cancer death rates')


```


## Quiz!


```{r quiz2, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center'}
library(maps)
kc8084 <- read.delim('data\\gd80to84.txt')
names(kc8084) <- names(kc8084)[-grep('x', tolower(names(kc8084)))]

kc8589 <- read.delim('data\\gd85to89.txt')

kc <- data.frame(state=kc8589$state, 
                 county=kc8589$county, 
                 aakc=kc8084$aadc + kc8589$aadc
                 #aakc=10^5*(kc8084$dc + kc8589$dc)/(kc8084$pop + kc8589$pop)
)

q90 <- with(kc, quantile(aakc, probs = 0.9, na.rm=T))
kc90 <- subset(kc, aakc>q90)

counties.90 <- paste0(tolower(kc90$state),',',tolower(kc90$county))


q10 <- with(kc, quantile(aakc, probs = 0.1, na.rm=T))
kc10 <- subset(kc, aakc<q10)

counties.10 <- paste0(tolower(kc10$state),',',tolower(kc10$county))

par(mfrow=c(1,2))
library(maps)
map('state')
map('county', counties.90, col='black', fill=T, add=T)
title('Highest kidney cancer death rates')

map('state')
map('county', counties.10, col='black', fill=T, add=T)
title('Lowest kidney cancer death rates')
```


## Quiz Answer

- Mix simulation of repeated i.i.d trials with 20% of success and sort

```{r, echo=FALSE, message=FALSE, warning=FALSE}

x.10 <- rbinom(10, 10, prob = 0.2)
x.100 <- rbinom(10, 100, prob = 0.2)
x.1000 <- rbinom(10, 1000, prob = 0.2)

dd <- data.frame(d = c(paste0(x.10,'/10'), paste0(x.100,'/100'), paste0(x.1000,'/1000')),
           rate = c(x.10/10, x.100/100, x.1000/1000))

output <- dd[sort.int(dd$rate, decreasing = FALSE, index.return = T)$ix,'d']
output <- as.character(output)

cat(output[1:5],'\n','\n', output[6:10],'\n','\n', output[11:15],'\n',
            '\n',output[16:20],'\n','\n', output[21:25],'\n', '\n',output[26:30])                 
```

> - More variation in observed rates with smaller sample sizes

> - _Answer:_ High cancer and low cancer counties are small populations

## Bayesian Mechanics
<!-- THIS IS A COMMENTED TEXT EVEN FOR MARKDOWN :-) -->

1. Set up full probability model
    - for all observable & unobservable quantities
    - consistent with problem knowledge & data collection
2.  Condition on observed data
    + calculate posterior probability of unobserved quantities conditional on observed quantities
3. Evaluate
    + model fit
    + implications of posterior


## Notation 

- Basic Quantities
    + $y$: observed data
    + $\theta$: parameters (and other unobserved quantities)
    + $x$: constants, predictors for conditional models
- Basic Predictive Quantities
    - $\tilde y$: unknown, potentially observable quantities
    - $\tilde x$: predictors for unknown quantities
    
## Naming Conventions

- __Joint:__ $p(y,\theta)$

- __Sampling/Likelihood:__ $p(y\vert \theta)$
    - Sampling is function of $y$ with $\theta$ fixed (probability function)
    - Likelihood is function of $\theta$ with $y$ fixed (_not_ probability function)
- __Prior:__ $p(\theta)$
- __Posterior:__ $p(\theta\vert y)$
- __Data Marginal (Evidence):__ $p(y)$
- __Posterior Predictive:__ $p(\tilde y\vert y)$

## Bayes's Rule for Posterior
\begin{equation}
\begin{array}{rcl}
p(\theta\vert y) &=& \frac{p(y, \theta)}{p(y)} \\
\\
 & = & \frac{p(y\vert\theta)p(\theta)}{p(y)} \\
\\
& = & \frac{p(y\vert\theta)p(\theta)}{\int_\Theta p(y\vert\theta)d\theta}\\
\\
& = & \frac{p(y\vert\theta)p(\theta)}{\int_\Theta p(y \vert\theta)p(\theta)d\theta}
\end{array}
\end{equation}

- _Inversion:_ Final result depends only on sampling distribution (likelihood) $p(y\vert\theta)$ and prior $p(\theta)$

##Bayes's Rule up to Proportion

- If data $y$ are fixed, then
$$
\begin{array}{rcl}
p(\theta\vert y ) & = & \frac{p(y\vert\theta)p(\theta)}{p(y)} \\
\\
& \propto & p(y\vert\theta)p(\theta) \\
\\
& = & p(y,\theta)
\end{array}
$$

- Posterior proportional to likelihood times prior
- Equivalently, posterior proportional to joint

## Posterior Predictive Distribution

- Predict new data $\tilde y$ based on observed data $y$
- Marginalize out parameter from posterior
$$
p(\tilde y \vert y) = \int_\Theta p(\tilde y \vert \theta) p(\theta \vert y)d\theta
$$
averaging predcitions $p(\tilde y\vert\theta)$ weighted by posterior $p(\theta\vert y)$
- $\Theta = \left\{\theta \vert p(\theta \vert y) >0 \right\}$ is the support of $p(\theta\vert y)$
- For discrete parameters $\theta$,
$$
p(\tilde y \vert y)=\sum_{\theta \in \Theta} p(\tilde y \vert \theta)p(\theta\vert y)
$$

- Can mix continuous and discrete (integral as shorthand)

## Event Probabilities

- Recall that an event A is a collection of outcomes 
- Suppose event $A$  is determined by indicator on parameters
$$
f(\theta) = \begin{cases} 1 & \mbox{if } \theta \in A \\ 
0 & \mbox{if } \theta \notin A  \end{cases}
$$
- e.g. $f(\theta) = \theta_1 > \theta_2$ for $Pr(\theta_1 > \theta_2\vert y)$
- Bayesian event probabilities compute posterior mass
$$
P(A) = \int_\Theta f(\theta)p(\theta\vert y)d\theta
$$
- Not frequentist, because involves parameter probabilities

## Laplace's Data and Problems
- Laplace's data on live birth in Paris from 1745-1770



_sex_ |_live births_
------|------------
female| 241,945
male  | 251,527

- Question 1 (Event Probability)
\par
Is a boy more likely to be born than a girl?

- Question 2 (Estimate)
\par
What is the birth rate of boys vs. girls?

- Bayes formulated the basic binomial model 

<!-- Laplace solved the integral (Bayes couldn't) -->

## Binomial Distribution
- Binomial distribution is number of successes $y$ in $N$ i.i.d. Bernoulli trials with chance of success $\theta$
- If $y_1,\ldots , y_N\sim$ Bernoulli($\theta$),
\par then $(y_1 + \ldots + y_N)\sim$ Binomial($N,\theta$)
- The probabilistic model is
$$
Binomial (y\vert N, \theta) = {{N}\choose{y}}\theta^y(1-\theta)^{N-y}
$$
where the binomial coefficient normalizes for permutation (i.e., which subset of $n$ has $y_n =1$):
$$
{{N}\choose{y}} = \frac{N!}{y!(N-y)!}
$$

## Bayes's Binomial Model

- Data
    - $y:$ total number of male live births (241,945)
    - $N:$ total number of live births (493,472)
- Parameter
    - $\theta \in (0,1)$: proportion of male live births
- Likelihood
$$
p(y\vert N,\theta) = Binomial(y\vert N,\theta) ={{N}\choose{y}}\theta^y(1-\theta)^{N-y}
$$
- Prior
$$
p(\theta) = Uniform(\theta\vert 0,1)=1
$$

## Beta Distribution
- Required for analytic posterior of Bayes's model
- For parameters $\alpha, \beta>0$ and $\theta\in (0,1)$,
$$
Beta(\theta\vert \alpha, \beta) = \frac{1}{B(\alpha,\beta)}\theta^{\alpha-1}(1-\theta)^{\beta-1}
$$
- Euler's Beta function is used to normalize
$$
Beta(\theta\vert \alpha, \beta) = \int_0^1 u^{\alpha-1}(1-u)^{\beta-1}du = \frac{\Gamma(\alpha)\Gamma(\beta)}{\Gamma(\alpha+\beta)}
$$
where $\Gamma()$ is continuous generalization of factorial
- Note: $Beta(\theta\vert 1,1)=Uniform(\theta, 0,1)$

## Beta Distribution examples

```{r,  echo=F, fig.width=10, fig.height=10}
suppressWarnings(library(LearnBayes))
pippo<-function (prior, data, where = "topright") 
{
    a = prior[1]
    b = prior[2]
    s = data[1]
    f = data[2]
    n=s+f
    a1<-a+s
    b1<-b+f
    p = seq(0.005, 0.995, length = 500)
    prior = dbeta(p, a, b)
    like = dbeta(p, s + 1, f + 1)
    #like = dbinom(s,n,p)
    post = dbeta(p, a + s, b + f)
    m = max(c(prior, like, post))
  plot(p, post, type = "l", ylab = "Density", xlab=expression(theta), lty = 1, lwd = 3, 
         main = "", ylim = c(0, m), col = "blue")
  #  lines(p, prior, lty = 3, lwd = 3, col = "green")
  #  lines(p, like, lty = 1, lwd = 3, col = "blue")
  #  legend(where, c("Prior", "Likelihood", "Posterior"), lty = c(3, 
  #      1, 2), lwd = c(3, 3, 3), col = c("green", "blue", "red"))
  text(0.9,dbeta(0.6, a+s,b+f), paste('n =',data[1]+data[2],'\ny =',data[1]))
}
par(mfrow=c(2,2))
prior=c(1, 1)  # proportion has a beta(2, 10) prior
data=c(3, 2)   # observe 10 successes and 30 failures
pippo(prior,data)

data=c(12, 8)   # observe 10 successes and 30 failures
pippo(prior,data)

data=c(60, 40)   # observe 10 successes and 30 failures
pippo(prior,data)

data=c(600, 400)   # observe 10 successes and 30 failures
pippo(prior,data)

```

## Posterior distribution
- Given Baye's general formula for the posterior
$$
p(\theta\vert y, N)=\frac{Binomial(y\vert N,\theta)Uniform(\theta\vert 0,1)}{\int_\Theta Binomial(y\vert N,\theta')p(\theta')d\theta'}
$$
- Using Euler's Beta function to normalize the posterior with final solution
$$
p(\theta\vert y, N)= Beta(\theta\vert y+1, N-y+1)
$$

## Estimation
- Posterior is $Beta(\theta \vert 1+ 241,945, 1+ 251,527)$
- Posterior mean:
$$
\frac{1+ 241,945}{1+ 241,945+1+ 251,527}\approx 0.4902913
$$
- Maximum likelihood estimate same as posterior mode (because of uniform prior)
$$
\frac{241,945}{241,945+251,527}\approx 0.4902912
$$
- As number of observations $\to \infty$, MLE approaches to posterior mean


## Event Probability Inference
- What is probability that a male live birth is more likely than a female live birth?

$$
\begin{array}{rcl}
Pr[\theta>0.5] & = & \int_\Theta I[\theta>0.5]p(\theta\vert y,N)d\theta \\
\\
& = & \int_{0.5}^1 p(\theta\vert y,N)d\theta \\
\\
& = & 1- F_{\theta\vert y,N}(0.5))\\
\\
&\approx & 1^{-42}
\end{array}
$$
- $I[\theta>0.5] =1$ if $\theta>0.5$ is true and 0 otherwise
- $F_{\theta\vert y,N}$ is posterior cumulative distribution function (cdf)


## Bayesian Fisher Exact Test

- Suppose we observe the following data on handedness

 _Drug_ | _AEs_      |_Successes_|  Total
------- |------------|-----------|------
_A_     |9$(y_1)$    | 43        | 52 $(N_1)$
_B_     | 4$(y_2)$   |44         | 48$(N_2)$



- Assume likelihoods $Binomial(y_k\vert N_k, \theta_k)$, uniform priors
- Is Drug B better than A?
$$
\begin{array}{rcl}
Pr[\theta_1 > \theta_2 \vert y, N] &=&
\int_0^1 \int_0^1 p(\theta_1\vert y_1, N_1)p(\theta_2\vert y_2, N_2)d\theta_1d\theta_2\\
 &\approx & 0.91
 \end{array}
$$

- Directly interpretable result; _not_ frequentist procedure

## Visualizing Posterior Difference
- Plot of posterior difference $p(\theta_2 -\theta_1 \vert y N)$ (B-A)

```{r, echo=FALSE}
library(LearnBayes)

suppressWarnings(library(LearnBayes))
pippo<-function (prior, data, where = "topright") 
{
    a = prior[1]
    b = prior[2]
    s = data[1]
    f = data[2]
    n=s+f
    a1<-a+s
    b1<-b+f
    p = seq(0.005, 0.995, length = 500)
    prior = dbeta(p, a, b)
    like = dbeta(p, s + 1, f + 1)
    #like = dbinom(s,n,p)
    post = dbeta(p, a + s, b + f)
    m = max(c(prior, like, post))
  plot(p, post, type = "l", ylab = "Density", xlab=expression(theta), lty = 1, lwd = 3, 
         main = "", ylim = c(0, m), col = "blue")
  #  lines(p, prior, lty = 3, lwd = 3, col = "green")
  #  lines(p, like, lty = 1, lwd = 3, col = "blue")
  #  legend(where, c("Prior", "Likelihood", "Posterior"), lty = c(3, 
  #      1, 2), lwd = c(3, 3, 3), col = c("green", "blue", "red"))
  text(0.9,dbeta(0.6, a+s,b+f)+1, paste('n =',data[1]+data[2],'\ny =',data[1]))
}
par(mfrow=c(1,2))
prior=c(1, 1)  # proportion has a beta(2, 10) prior
dataA=c(43, 9)   # observe 10 successes and 30 failures
dataB = c(44, 4)
pippo(prior,dataA)
pippo(prior,dataB)


```


## Visualizing Posterior Difference
- Plot of posterior difference $p(\theta_2 -\theta_1 \vert y N)$ (B-A)

```{r, echo=FALSE}
library(LearnBayes)

suppressWarnings(library(LearnBayes))
pippo<-function (prior, data, where = "topright") 
{
    a = prior[1]
    b = prior[2]
    s = data[1]
    f = data[2]
    n=s+f
    a1<-a+s
    b1<-b+f
    p = seq(0.005, 0.995, length = 500)
    prior = dbeta(p, a, b)
    like = dbeta(p, s + 1, f + 1)
    #like = dbinom(s,n,p)
    post = dbeta(p, a + s, b + f)
    m = max(c(prior, like, post))
  plot(p, post, type = "l", ylab = "Density", xlab=expression(theta), lty = 1, lwd = 3, 
         main = "", ylim = c(0, m), col = "blue")
  #  lines(p, prior, lty = 3, lwd = 3, col = "green")
  #  lines(p, like, lty = 1, lwd = 3, col = "blue")
  #  legend(where, c("Prior", "Likelihood", "Posterior"), lty = c(3, 
  #      1, 2), lwd = c(3, 3, 3), col = c("green", "blue", "red"))
  text(0.9,dbeta(0.6, a+s,b+f)+1, paste('n =',data[1]+data[2],'\ny =',data[1]))
}
par(mfrow=c(1,2))
prior=c(1, 1)  # proportion has a beta(2, 10) prior
dataA=c(43, 9)   # observe 10 successes and 30 failures
dataB = c(44, 4)
pippo(prior,dataA)
pippo(prior,dataB)


```

## Visualizing Posterior Difference {.smaller}

```{r, bugs, cache=T}

library(R2OpenBUGS)
mymodel = function(){
  y1~ dbin(theta1, N1)
  y2~ dbin(theta2, N2)
  theta1~dbeta(1,1)
  theta2~dbeta(1,1)
  theta <- theta2-theta1
}

data = list(N1 =52, N2 =48, y1=43, y2 = 44)
parameters=c("theta")
bugsInits <- function() {
  return(list(theta1=0.5, theta2=0.5)) 
}

resbugs = bugs(data, inits=bugsInits, parameters.to.save=parameters, model.file=mymodel,
n.chains=1, n.iter=10000, n.burnin=2000, n.thin=1,
                debug=TRUE)


```

## Visualizing Posterior Difference

- Plot of posterior difference $p(\theta_2 -\theta_1 \vert y N)$ (B-A)
- compute 95% credible interval

```{r dependson="bugs", fig.height=3, fig.width=3}
theta.post = resbugs$sims.list$theta
plot(density(theta.post), main= "posterior probability", 
     xlab=expression(theta[2]-theta[1]) )
quantile(theta.post, probs=c(.025, .5, .975))

```

## Visualizing Posterior Difference

- Plot of posterior difference $p(\theta_2 -\theta_1 \vert y N)$ (B-A)
- compute 95% credible interval

```{r dependson="bugs", fig.height=3, fig.width=3}
theta.post = resbugs$sims.list$theta
quantile(theta.post, probs=c(.025, .5, .975))

```


## Conjugate Priors
- Family $F$ is a conjugate prior for family $G$ if
    - prior in $F$ and
    - likelihood in $G$
    - entails posterior in $F$

- Before MCMC techniques became practical, Bayesian analysis mostly involved conjugate priors
- Still widely used because analytic solutions are more efficient than MCMC

## Beta is Conjugate to Binomial

\begin{itemize}
\item \begin{tabular}{p{2cm}p{1.6cm}p{2cm}} Prior & $p(\theta \vert \alpha, \beta) =$ & $Beta(\theta\vert\alpha, \beta)$ \end{tabular}

\item \begin{tabular}{p{2cm}p{1.6cm}p{2cm}} Likelihood & $p(y \vert N, \theta) =$ & $Binomial(y\vert N, \theta)$ \end{tabular}
\item \begin{tabular}{p{2cm}p{1.6cm}p{2cm}} 
Posterior &  & \end{tabular}
\end{itemize}
\vspace{-1cm}
$$
\begin{array}{rcl}
p(\theta\vert y, N, \alpha,\beta) &\approx& p(\theta\vert\alpha,\beta)p(y\vert N,\theta)\\
\\
&=& Beta(\theta\vert\alpha,\beta)Binomial(y\vert N,\theta)\\
\\
&=& \frac{1}{B(\alpha, \beta)}\theta^{\alpha-1}(1-\theta)^{\beta-1}{{N}\choose{y}}\theta^y(1-\theta)^{N-y}\\
\\
&\approx&\theta^{y+\alpha-1}(1-\theta)^{N-y+\beta-1}\\
\\
&\approx&Beta(\theta\vert \alpha + y, \beta +(N-y))
\end{array}
$$

## Chaining Updates

- Start with prior $Beta(\theta\vert\alpha,\beta)$
- Receive binomial data in $K$ stages $(y_1, N_1),\ldots , (y_K, N_K)$
- After $(y_1, N_1)$, posterior is $Beta(\theta\vert \alpha + y_1, \beta +N_1-y_1)$
- Use as prior for $(y_2, N_2)$ with posterior
$$
Beta(\theta\vert \alpha + y_1 +y_2, \beta +(N_1-y_1) + (N_2-y_2))
$$
- Lather, rinse, repeate until final posterior
$$
Beta(\theta\vert \alpha + y_1 +\ldots +y_K, \beta +(N_1 +\ldots N_K) - (y_1+\ldots +y_K))
$$
- Same results as if we had updated with combined data
$$
Beta(y_1+\ldots +y_K, N_1 +\ldots N_K)
$$

## MAP Estimator 

- For a Bayesian model $p(y,\theta) = p(y\vert\theta)p(\theta)$, the max a posteriori (MAP) estimate maximizes the posterior
$$
  \begin{array}{rcl}
\theta^* & = & arg max_\theta p(\theta\vert y)\\
\\
& = & arg max_\theta \frac{p(y\vert \theta)p(\theta)}{p(y)}\\
\\
& = & arg max_\theta p(y\vert \theta)p(\theta)\\
\\
& = & arg max_\theta \log p(y\vert \theta) + \log p(\theta)\\
\end{array}
$$
  - _not_ Bayesian because it does not integrate over uncertainty
- _not_ frequentist because of distributions over parameters

## MAP and the MLE
- MAP estimate reduces to the MLE if the prior is uniform, i.e.,
$$
  p(\theta) = c
  $$
    because
  $$
    \begin{array}{rcl}
  \theta^* & = & arg max_\theta p(y\vert \theta)p(\theta)\\
  \\
  & = & arg max_\theta p(y\vert \theta)c\\
  \\
  & = & arg max_\theta p(y\vert \theta)\\
  \end{array}
  $$
  

## Penalized Maximum Likelihood
  - The MAP can be made palatable to frequentists via philosophical sleight of hand
  - Treat the negative log prior $-\log p(\theta)$ as a _penalty_
  - e.g. a $Normal(\theta\vert\mu,\sigma)$ prior becomes a penalty function
  $$
    \lambda_{\theta,\mu,\sigma} = - \left(
      \log\sigma + \frac{1}{2}\left(\frac{\theta-\mu}{\sigma}\right)^2\right)
    $$
      - Maximize sum of log likelihood and negative penalty
    
$$
  \begin{array}{rcl}
  \theta^* & = & arg max_\theta \log p(y\vert \theta, x) - \lambda_{\theta,\mu,\sigma}\\
    \\
  & = & arg max_\theta \log p(y\vert \theta, x) + \log p(\theta\vert\mu,\sigma)
  \end{array}
  $$
  
## Proper Bayesian Point Estimates

- Choose estimate to minimize some loss function

- To minimize expected squared error (L2 loss), $E[(\theta - \theta^{'})^2]$, use the posterior mean

$$
\hat\theta = arg min_\theta E[(\theta - \theta^{'})^2] = \int_\Theta \theta\times p(\theta\vert y)d\theta
$$

- To minimize expected absolute error (L1 loss), $E[|\theta - \theta^{'}|]$, use posterior median

- Other loss (utility) functions possible, study of which falls under decision theory

- All share property of involving full Bayesian inference.


## Getting the slides

* The slides for this course were created with Rmarkdown: [http://rmarkdown.rstudio.com/](http://rmarkdown.rstudio.com/). 
* They are available from [https://github.com/berkeley3/BDA](https://github.com/berkeley3/BDA).
* To re-compile the slides:

    + Download the directory containing the lectures from Github
    + In R open the .Rmd file and set the working directory to the lecture directory
    + Click the *KnitHTML* button on Rstudio or run the following commands: 
  
```{r RmarkdownChunk, eval=FALSE}
library(rmarkdown) 
render("main.Rmd")
```